---
title: "R Case Study 1"
author: "Vibha Srinivasa Vibha Srinivasa"
date:  
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: simplex
    highlight: zenburn
---

##### To complete this assignment, follow these steps:

1. Download the `Rcasestudy1.Rmd` file from Canvas.

2. Open `Rcasestudy1.Rmd` in RStudio.

3. Replace the "Your Name Here" text in the `author:` field with your own name.

4. Supply your solutions to the homework by editing `Rcasestudy1.Rmd`.

5. When you have completed the homework and have **checked** that your code runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `Rcasestudy1_YourNameHere.Rmd`, and submit on Canvas.  (YourNameHere should be changed to your own name.)

##### Homework tips:

1. Recall the following useful RStudio hotkeys.

|Keystroke      | Description |
|----------------|-----------------------------------------------------------------------------|
| `<tab>`        | Autocompletes commands and filenames,  and lists arguments for functions.|
| `<up>`         | Cycles through previous commands in the console prompt |
| `<ctrl-up>`    | Lists history of previous commands matching an unfinished one |
| `<ctrl-enter>` | Runs current line from source window to Console. Good for trying things out ideas from a source file. |
| `<ESC>`        | Aborts an unfinished command and get out of the + prompt |




**Note**: Shown above are the Windows/Linux keys.  For Mac OS X, the `<ctrl>` key should be substituted with the `<command>` (&#8984;) key.

2. Instead of sending code line-by-line with `<ctrl-enter>`, you can send entire code chunks, and even run all of the code chunks in your .Rmd file. Look under the <Chunks> menu of the Source panel.

3. Run your code in the Console and Knit HTML frequently to check for errors.

4. You may find it easier to solve a problem by interacting only with the Console at first. 

### Problem 1: Vector Boolean operations (15 points)

##### R has two kinds of Boolean operators implemented, single (`&`, `|`) and double (`&&`, `||`).  

One of these operators takes advantage of something called *lazy evaluation* while the other does not.  They also do not behave the same way when applied to *vectors*.  

Read the help file (`help("||")`) and construct some examples to help figure out how the two behave.  

To help you get started, try out the following two examples in your console:

```{r, eval = FALSE}
# Example:  The variable y.prob2a is never defined.  
# (Do not define it!)
# What happens when you run this code?
x.prob2a <- 5
(x.prob2a < 10) | (y.prob2a > 2)
(x.prob2a < 10) || (y.prob2a > 2)
```

```{r, eval = FALSE}
# Define vectors
x.prob2a.vec <- c(TRUE, FALSE, FALSE)
y.prob2a.vec <- c(TRUE, TRUE, FALSE)

# Apply various Boolean operations to see what happens
x.prob2a.vec & y.prob2a.vec
x.prob2a.vec && y.prob2a.vec
x.prob2a.vec | y.prob2a.vec
x.prob2a.vec || y.prob2a.vec
```

Can you explain what's happening?  Write up a brief explanation below.

<font color="#336600">

**Replace this text with your explanation.**

</font>


### Problem 2: Indego Bike Share Service (25 points)

The questions below are to be answered on the data file Indego_trips_Q4_2016.csv, which is posted on Canvas. This file has been edited from the data provided by Indego at https://www.rideindego.com/about/data/
Indego is the bike share service in the City of Philadelphia, and the data file is a listing of every rental from every station for the fourth quarter of 2016. If you are not familiar with this valuable and popular service (there are over 175K rentals in the data file!), I suggest that you read more about it online.  As a part of the City’s commitment to sharing data, usage information on Indego is provided to residents regularly.  

(a) Create a data frame called Indego.df and read the contents of the csv file into it.
```{r}
# Edit me
Indego.df <- read_csv("C:\\Users\\vibhc\\Downloads\\Indego_trips_Q4_2016-1.csv")

```

(b)	Get summary information on this data frame.  What are the different pieces of information provided in the file?  How many trips are there in total?  How many of these trips are One Way?  How many are Round Trip?  
```{r}
# Edit me
summary(Indego.df)
```

(c) What are the different types of users?(Check passholder_type)  What percentage of the users had the Indego30 pass?  What percentage had the IndegoFlex pass?  What percentage were Walk-up customers?

```{r}
# Edit me
col <- Indego.df[12]
count(col)
print(sum(data$col3 == 'manoj'))
  
```

(d) Look at trip #5.  How long did this trip last?  (The duration data is given in seconds.)  What is the station ID where the trip started?  What is the station ID where the trip ended?  Using the station list provided on the above website, can you find the actual address where the trip started and ended?
```{r}
# Retrieve details for trip #5
trip5 <- Indego.df[5, ]

# Duration in seconds
trip5_duration <- trip5$duration

# Starting and ending station IDs
start_station_id <- trip5$start_station
end_station_id <- trip5$end_station

# Output trip #5 details
cat("Trip Duration (seconds):", trip5_duration, "\n")
cat("Start Station ID:", start_station_id, "\n")
cat("End Station ID:", end_station_id, "\n")

```

(e) If you look at the summary information for this data frame, you will see that there are mins, means, etc., for the ID number of the starting station and the ID number of the ending station. However, the mean ID number is not a proper quantity. Convert both of these ID vectors into factors. Then, rerun the summary() function and report on the top 5 most frequently used starting and ending stations and how many trips originated or ended at each one.

```{r}
# Convert start_station and end_station to factors
Indego.df$start_station <- as.factor(Indego.df$start_station)
Indego.df$end_station <- as.factor(Indego.df$end_station)

# Rerun summary
summary(Indego.df)

# Count the most frequently used starting stations
start_station_counts <- table(Indego.df$start_station)
top_start_stations <- sort(start_station_counts, decreasing = TRUE)[1:5]

# Count the most frequently used ending stations
end_station_counts <- table(Indego.df$end_station)
top_end_stations <- sort(end_station_counts, decreasing = TRUE)[1:5]

# Output top 5 stations
cat("Top 5 Starting Stations:\n")
print(top_start_stations)

cat("\nTop 5 Ending Stations:\n")
print(top_end_stations)

```

(f) Look up the help file for the plot() function.  Using the information that you just learned, plot the latitude vs. longitude for the starting point of each trip (start_lat and start_lon).

```{r}
# Plot latitude vs. longitude of starting points
plot(
  Indego.df$start_lon,
  Indego.df$start_lat,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Starting Points of Trips",
  col = "blue",
  pch = 20
)

```

(g) Set a vector t equal to the trip duration.  Then, set tmin equal to t in minutes and thr equal to t in hours.

```{r}
# Create vector t for trip duration
t <- Indego.df$duration

# Convert duration to minutes and hours
tmin <- t / 60
thr <- t / 3600

# Add these columns to the data frame
Indego.df$tmin <- tmin
Indego.df$thr <- thr

# Output a preview of the updated data frame
head(Indego.df[, c("duration", "tmin", "thr")])

```

(h) Create a new data frame called indegomktg.df which includes the trip durations in hours and the passholder type.  Output its summary information.

```{r}
# Create indegomktg.df
indegomktg.df <- Indego.df[, c("thr", "passholder_type")]

# Summary of the new data frame
summary(indegomktg.df)

```

### Problem 3 (10 points)
(a)	Create a vector called shortpass that includes the passholder types for all the trips that lasted less than 1 hour.  What percentage of the short trips were completed by Indego30 pass holders?
```{r}
# Filter trips with duration less than 1 hour
short_trips <- Indego.df[Indego.df$thr < 1, ]

# Calculate percentage of short trips by Indego30 pass holders
shortpass <- short_trips$passholder_type
indego30_percentage <- sum(shortpass == "Indego30") / length(shortpass) * 100

cat("Percentage of short trips by Indego30 pass holders:", round(indego30_percentage, 2), "%\n")

```
(b)	Assume that you are looking to market podcasts and audiobooks stored on the bikes themselves (disposable earphones are also available). The podcasts tend to be around 30-40 minutes in length, whereas an audiobook is generally around 2-3 hours. There will be short ads in each file to generate revenue for Indego. To increase the customers' chances to listen to the offerings (and therefore the ads), figure out an excellent way to market podcasts or audiobooks at the beginning of the trip based on the passholder type.
```{r}
# Identify trips suitable for podcasts
podcast_trips <- Indego.df[Indego.df$thr >= 0.5 & Indego.df$thr <= 0.75, ]
podcast_passholder_types <- table(podcast_trips$passholder_type)

# Identify trips suitable for audiobooks
audiobook_trips <- Indego.df[Indego.df$thr >= 2 & Indego.df$thr <= 3, ]
audiobook_passholder_types <- table(audiobook_trips$passholder_type)

# Output marketing analysis
cat("Podcast Trips by Passholder Type:\n")
print(podcast_passholder_types)

cat("\nAudiobook Trips by Passholder Type:\n")
print(audiobook_passholder_types)

```

